Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-11 _tex/paper2.tex
Built site for gh-pages

\newcommand{\dsep}{\:\bot\:}
\newcommand{\ndsep}{\:\not\bot\:}
\newcommand{\cond}{\:|\:}

@@

Before incorporating the first theoretical principles of Thurstone's
theory, it is essential to further define SCMs. SCMs are formal
mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly \citep{Pearl_2009}.
Lastly, the functions, referred to as \emph{structural equations},
express the endogenous variables as non-parametric functions of other
endogenous and exogenous variables. These functions use the symbol
`\(:=\)' to denote the asymmetrical causal dependence between variables
and the symbol `\(\:\bot\:\)' to represent \emph{d-separation}, a
concept akin to (conditional) independence.

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model (see equation
(\ref{eq-mat27})). The vector contains \(n\) elements corresponding to
the number of rows in the \(R\) matrix, with each element
\(c_{(i,a,h,b,j,k)}\) being a binary value indicating the presence or
absence of data rows in \(R\), a definition similar to that of \(S\) in
equation (\ref{eq-mat26}).
\begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

The DAG~\ref{fig-cj14_dag} also integrates the comparison mechanism
\(C\) into the conceptual-population model, which ultimately derives the
final comparative outcome \(O_{R}\). First, it shows the
sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
that researchers cannot directly access this outcome due to the
comparison mechanism. Second, it depicts the \emph{comparison mechanism}
\(C\) as a causal factor influencing the final comparative outcome
\(O_{R}\). As with the sampling mechanism, a square encloses \(C\) to
indicate that it is a conditioned variable--that is, one that determines
\emph{which repeated judgments the subset of judges makes for the subset
of stimuli produced by the sampled individuals}. In essence, \(C\)
reflects the assumption that judges \emph{do not} perform all possible
repeated judgments for the selected stimuli. Instead, they carry out a
``sufficient number of comparisons'' \(n_{C}\) to allow for accurate
estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

Notably, DAG~\ref{fig-cj14_dag} shows that \(C\) is independent of all
other variables in the model. This independence implies that the
conceptual model represented by the DAG applies exclusively to Random
Allocation Comparative Designs \citep{Bramley_2015} or Incomplete Block
Designs \citep{Lawson_2015}. In these designs, every repeated judgment
has an equal probability of being included in the sample.

@@

Finally, without loss of generality, researchers can reformulate the
sample-comparison model in Figure~\ref{fig-cj14} into the equivalent
form shown in Figure~\ref{fig-cj15}. This reformulation rests on two
principles. First, because the sampling and comparison mechanism vectors
\(S\) and \(C\) are binary, researchers can integrate them into the
model using element-wise products. This approach eliminates the need to
define the structural equations \(f_{S}\) and \(f_{C}\). Second, since
only the final comparative outcome \(O_{R}\) is observed, it is standard
to assume that the distribution of the conceptual-population outcome
\(O^{cp}_{R}\), defined by the structural equation \(f_{O}\), also
applies to this outcome. This assumption removes the necessity of
describing the unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\).

In summary, extending Thurstone's general form allows researchers to
address issues with the BTL model, such as accounting for judge biases
(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}), and measurement error in
trait estimation and hypothesis testing (see
Section~\ref{sec-theory-issue2}). However, it does not address concerns
about stimuli' equal dispersions discussed in
Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
assumptions regarding the discriminal process distribution, it requires
a statistical model, which we develop in the next section.

\section{Abandoning the BTL model}\label{sec-statistical}

Researchers can easily derive a statistical model to address violations
of the equal dispersion assumption in CJ experiments (see
Section~\ref{sec-theory-issue1a}) using only SCM~\ref{fig-cj15_scm}.
This derivation is possible because the structural equations \(F\)
encode probabilistic information \citep{Pearl_et_al_2016}, which
researchers can replace with suitable probabilistic assumptions. As a
starting point, given the dichotomous nature of CJ outcomes,
model~\ref{fig-cj16_stat} assumes that \(O_{R}\) follows a Bernoulli
distribution\footnote{The binomial distribution--including its special
  case, the Bernoulli distribution-- represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that when only two un-ordered outcomes are possible, and the
  expected numbers of each type of event are assumed to be constant,
  then the binomial distribution is the one that is most consistent with
  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}.

Next, following the conventions of Generalized Linear Latent And Mixed
Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
model links the observed outcome to the latent discriminal difference
vector \(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Notice that the sampling
and comparison mechanisms, \(S\) and \(C\), act as element-wise
multipliers of \(D_{R}\) (denoted by the symbol \(\odot\)), directly
incorporating key assessment design features relevant to CJ experiments
into the statistical formulation. Researchers can omit these vectors
when only the observed CJ sample is available.

@@

@@

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} \odot S \odot C ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
\]

@@

\subcaption{\label{fig-cj16_scm}SCM~\ref{fig-cj15_scm}}

@@

\caption{\label{fig-cj16}Comparative judgment model, SCM and statistical
model assuming different discriminal dispersions for the student's
traits}

@@

Then, model~\ref{fig-cj16_stat} defines \(D_{R}\) as the difference
between the discriminal processes of two texts, \(T_{IA}[i, a]\) and
\(T_{IA}[h, b]\), which represent their underlying written-quality
trait. This difference also includes the corresponding repeated judge
biases, \(B_{JK}[j, k]\). Note that if we assume that \(B_{JK}[j,k]\)
reflects the difference in stimulus-specific biases, i.e.,
\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), then we can
re-write the discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] ] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). For clarity, the
square brackets in \(D_{R}\) only indicate the relevant indices within
each trait vector; they do not represent any data subsetting.

Next, the model defines \(T_{IA}\) as a linear combination of the
students' underlying writing-quality traits \(T_{I}\), the effects of
relevant text-related variables \(\beta_{XA}X_{IA}\) (e.g., text
length), and the texts' idiosyncratic errors \(e_{IA}\). The model
assumes that each element of \(e_{IA}\) is independent and identically
distributed \((iid)\) according to a Normal distribution, with a mean of
zero and a standard deviation \(p_{IA}\), defined as a proportion of
\(1\). This standardization establishes the scale of the latent trait,
as it is required by latent variable models \citep{Depaoli_2021}.

Additionally, the model defines \(T_{I}\) as a linear combination of the
effects of relevant student-related variables, \(\beta_{XI} X_{I}\), and
students' idiosyncratic errors, \(e_{I}\). As before, each element of
\(e_{I}\) is \(iid\) following a Normal distribution with a mean of
zero. However, the standard deviation, \(s_{XI}\), varies depending on
the teaching method to which each student belongs. Given the example in
Section~\ref{sec-theoretical}, where the teaching method
\(X_{I} = 1,2\), the model sets the following constraint to anchor the
scale of the latent trait \citep{Depaoli_2021}:
\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\) This constraint effectively addresses
concerns about the equal dispersion assumption raised in
Section~\ref{sec-theory-issue1a}.

Similarly, the model defines \(B_{JK}\) as a linear combination of the
judges' individual bias \(B_{J}\), the effects of relevant
judgment-related variables \(\beta_{ZK}Z_{JK}\) (e.g., the number of
judgments a judge makes), and judgment-specific idiosyncratic errors
\(e_{JK}\). The model assumes that each element of \(e_{JK}\) is \(iid\)
according to a Normal distribution, with a mean of zero and a standard
deviation \(p_{JK}\), defined as a proportion of \(1\), to anchor the
scale of the latent trait \citep{Depaoli_2021}.

The model also defines \(B_{J}\) as a linear combination of the effects
of relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
element of \(e_{J}\) is likewise assumed to be \(iid\) Normal with mean
zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
which each judge belongs. For instance, if \(Z_{J} = 1, \dots, 3\)
represents three groups of judges with varying expertise, the model
imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
scale of the latent trait \citep{Depaoli_2021}.

Finally, we apply \emph{Bayesian inference methods} to turn
model~\ref{fig-cj16_stat} into a practical statistical tool for
analyzing paired comparison data. The \textbf{Declarations} section of
this document provides a link to this model, along with an alternative
specification that assumes equal discriminal dispersions. We implemented
both models in \texttt{Stan} \citep[version \(2.26.1\),][]{Stan_2020}.

@@

\subsection{Findings}\label{sec-discussion-finding}

@@

\begin{equation}\phantomsection\label{eq-group_diff}{
E[Y_{i} \:|\:T_{i}=1] - E[Y_{i} \:|\:T_{i}=2]
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-ICE}{
\tau_{i} = Y_{i} \:|\:do(T_{i}=1) - Y_{i} \:|\:do(T_{i}=2)
}\end{equation}

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \:|\:do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \:|\:do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address this challenge, the structural approach extends
the ICE to the \emph{average causal effect} (ACE,
Equation~\ref{eq-ACE}), representing the average difference between the
students' observed potential outcomes and their counterfactual
counterparts.

\begin{equation}\phantomsection\label{eq-ACE}{
\begin{aligned}
\tau & = E[\tau_{i}] \\
  & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)]
\end{aligned}
}\end{equation}

Even though counterfactuals are unobservable, researchers can still
identify the ACE from associational estimates by leveraging the
structural approach. The approach identifies the ACE by statistically
conditioning data on a \emph{sufficient adjustment set} of variables
\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. When such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \:|\:X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).

@@

\begin{equation}\phantomsection\label{eq-CACE}{
CACE_{t} = E[Y_{i} \:|\:T_{i}=t, X]
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mACE1}{
\begin{aligned}
  \tau & = E[Y_{i} \:|\:do(T_{i}=1)]- E[Y_{i} \:|\:do(T_{i}=2)] \\
  & = E_{X}[CACE_{1} - CACE_{2}] \\
  & = E_{X}\left[ E[Y_{i} \:|\:T_{i}=1, X] - E[Y_{i} \:|\:T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-prop_rule}{
P(\theta \:|\:X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\:|\:\theta) \cdot P(\theta)
}\end{equation}

>>>>

\newcommand{\dsep}{\:\bot\:}
\newcommand{\ndsep}{\:\not\bot\:}

@@

Before incorporating the first theoretical principles of Thurstone's
theory, it is essential to further define SCMs. SCMs are formal
mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\)
\citep{Pearl_2009, Pearl_et_al_2016, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly \citep{Pearl_2009}.
Lastly, the functions, referred to as \emph{structural equations},
express the endogenous variables as non-parametric functions of other
endogenous and exogenous variables. These functions use the symbol
`\(:=\)' to denote the asymmetrical causal dependence between variables
and the symbol `\(\:\bot\:\)' to represent \emph{d-separation}, a
concept akin to (conditional) independence.

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model (see equation
(\ref{eq-mat27})). The vector contains \(n\) elements corresponding to
the number of rows in the \(R\) matrix, with each element
\(c_{(i,a,h,b,j,k)}\) being a binary value indicating the presence or
absence of data rows in \(R\), a definition similar to that of
\(s_{(i,a,h,b,j,k)}\) in equation (\ref{eq-mat26}).
\begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

The DAG~\ref{fig-cj14_dag} also integrates the \emph{comparison
mechanism} \(C\) into the conceptual-population model, ultimately
deriving the outcome \(O_{R}\). It shows the sample-comparison outcome
\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
directly access this outcome due to the comparison mechanism. Moreover,
it depicts \(C\) as a causal factor influencing \(O_{R}\). As with the
sampling mechanism, a square encloses \(C\) to indicate that it is a
conditioned variable--that is, one that determines \emph{which repeated
judgments the subset of judges makes for the subset of stimuli produced
by the sampled individuals}. In essence, \(C\) reflects the assumption
that judges \emph{do not} perform all possible repeated judgments for
the selected stimuli. Instead, they carry out a ``sufficient number of
comparisons'' \(n_{C}\) to allow for accurate estimation of the
proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

Notably, DAG~\ref{fig-cj14_dag} also shows that \(C\) is independent of
all other variables in the model. This independence implies that the
conceptual model represented by the DAG applies exclusively to Random
Allocation Comparative Designs \citep{Bramley_2015} or Incomplete Block
Designs \citep{Lawson_2015}. In these designs, every repeated judgment
has an equal probability of being included in the sample.

@@

Finally, since it is standard to assume that the distribution of the
conceptual-population outcome \(O^{cp}_{R}\) also holds for
\(O^{sc}_{R}\) and \(O_{R}\), we can reformulate the sample-comparison
model in Figure~\ref{fig-cj14} into the equivalent form shown in
Figure~\ref{fig-cj15}. This reformulation produces a model that applies
directly to a sample of comparative data. In this version, the
unobserved outcomes \(O^{cp}_{R}\) and \(O^{sc}_{R}\) are omitted, and
\(O_{R}\) inherits the structural equation \(f_{O}\) that originally
defined \(O^{cp}_{R}\). Moreover, the definition of \(O_{R}\) now
reflects its direct dependence on the discriminal difference \(D_{R}\)
and the sample and comparison mechanisms, \(S\) and \(C\).

In summary, the SCM~\ref{fig-cj15_scm} and DAG~\ref{fig-cj15_dag} extend
Thurstone's general form to address several limitations of the BTL
model. These extensions account for judge biases (see
Section~\ref{sec-theory-issue1b}), reflect the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}), and incorporate
measurement error in trait estimation and hypothesis testing (see
Section~\ref{sec-theory-issue2}). However, they do not resolve concerns
about the assumption of equal dispersions among stimuli discussed in
Section~\ref{sec-theory-issue1a}. Because this issue involves
statistical assumptions about the distribution of the discriminal
process, it requires a statistical model, which we develop in the next
section.

\section{From SCMs to a statistical model}\label{sec-statistical}

With the structure in SCM~\ref{fig-cj15_scm}, we can derive a
statistical model that addresses violations of the equal dispersion
assumption in CJ experiments (see Section~\ref{sec-theory-issue1a}).
This derivation is possible because a fully specified SCM encodes
functional and probabilistic information, which we can replace with
appropriate functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Specifically, SCM~\ref{fig-cj15_scm} allows us
to express the joint distribution of our CJ system as a product of
simpler conditional probability distributions (CPDs)\footnote{This
  re-expression is possible because the \emph{chain rule} of probability
  and the \emph{Bayesian Network Factorization (BNF)} property. For
  further details, see \citet{Pearl_et_al_2016} and \citet{Neal_2020}.},
as shown in equation (\ref{eq-mat51}). For clarity, we treat expressions
such as \(Y := f_{Y}(X)\), \(P(Y \mid X)\), and \(Y \sim f(Y \mid X)\)
as equivalent. Here, \(P(Y \mid X)\) and \(f(Y \mid X)\) represent the
CPD of \(Y\) given \(X\).
\begin{equation}\phantomsection\label{eq-mat51}{
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
}\end{equation}

Each CPD in equation (\ref{eq-mat51}) rests on specific assumptions,
which we outline below. First, we assume that \(O_{R}\) follows a
Bernoulli distribution\footnote{The binomial distribution--including its
  special case, the Bernoulli distribution-- represent a maximum entropy
  distribution for binary events \citep[pp.~34]{McElreath_2020}. This
  means that when only two un-ordered outcomes are possible, and the
  expected numbers of each type of event are assumed to be constant,
  then the binomial distribution is the one that is most consistent with
  these constraints \citep[pp.~310]{McElreath_2020}. For a detailed
  discussion of the binomial as a maximum entropy distribution, see
  \citet[chap.~10.1.2]{McElreath_2020}.}, reflecting the binary nature
of CJ outcomes. Furthermore, following the conventions of Generalized
Linear Latent and Mixed Models (GLLAMMs)
\citep{Rabe_et_al_2004a, Skrondal_et_al_2004a, Rabe_et_al_2012b}, the
distribution links \(O_{R}\) to the latent discriminal difference vector
\(D_{R}\) using an inverse-logit function:
\(\text{inv\_logit}(x) = 1/(1 + \exp(-x))\). Note that the probability
distributions for the sample and comparison mechanisms, \(P(S)\) and
\(P(C)\), included in the probabilistic model \ref{fig-cj16_prob}, are
omitted from the statistical model \ref{fig-cj16_stat}. We exclude these
probabilities from the statistical definition because, although they are
part of the joint distribution of the data, they do not influence the
dependence between \(O_{R}\) and \(D_{R}\), since \(S\) and \(C\) are
governed by random selection designs\footnote{Randomization ensures that
  data---and, by extension, an estimator---satisfies several key
  identification properties, such as common support, no interference,
  and consistency. The most critical property, however, is the
  elimination of confounding. \emph{Confounding} occurs when an external
  variable, such as \(X_{I}\), simultaneously influences both the
  outcome (e.g., \(O_{R}\)) and a variable of interest (e.g., \(S\)),
  resulting in spurious associations between the two
  \citep{Everitt_et_al_2010}. Randomization addresses this issue by
  effectively decoupling the association between the variable of
  interest and any other variable, except for the outcome itself. For a
  more detailed discussion on the benefits of randomization, see
  \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, \citet{Neal_2020}, and
  \citet{Hernan_et_al_2020}.}.

Second, we define \(D_{R}\) as the difference between the discriminal
processes of two texts, \(T_{IA}[i, a]\) and \(T_{IA}[h, b]\), plus the
corresponding repeated judge biases, \(B_{JK}[j, k]\). The discriminal
processes represent the underlying written-quality traits of the texts.
Note that if we assume that \(B_{JK}[j,k]\) reflects the difference in
stimulus-specific biases, i.e.,
\(B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]\), we can re-write the
discriminal difference as:
\begin{equation}\phantomsection\label{eq-mat52}{
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
}\end{equation} This formulation reveals that the discriminal difference
captures a \emph{pure interaction effect}, in which neither the texts'
discriminal processes nor the judges' biases alone determine the
outcome, but their interaction does \citep{Attia_et_al_2022}. In simple
terms, the discriminal processes of the stimuli become an observable
outcome only through the judges' perceptions (biases). For clarity, the
square brackets in \(D_{R}\) only indicate the relevant indices within
each trait vector; they do not represent any data subsetting.

@@

\begin{minipage}{0.33\linewidth}

@@

\[
\begin{aligned}
  O_{R} & := f_{O}(D_{R}, S, C) \\ 
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
  e_{IA} & \:\bot\:e_{JK} \\ \\
\end{aligned}
\]

\subcaption{\label{fig-cj16_scm}SCM}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\[
\begin{aligned}
  & P( O_{R} \mid D_{R}, S, C ) \cdot P(S) \cdot P(C) \\
  & P( D_{R} \mid T_{IA}, B_{JK} ) \\
  & P( T_{IA} \mid T_{I}, X_{IA}, e_{IA} ) \\
  & P( T_{I} \mid X_{I}, e_{I} ) \\
  & P( B_{JK} \mid B_{J}, Z_{JK}, e_{JK} ) \\
  & P( B_{J} \mid Z_{J}, e_{J} ) \\
  & P( e_{I} ) \\
  & P( e_{J} ) \\
  & P( e_{IA} ) \\  
  & P( e_{JK} )
\end{aligned}
\]

\subcaption{\label{fig-cj16_prob}Probabilistic model}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\[
\begin{aligned}
  O_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,s_{XI}) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,s_{ZJ}) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
\]

@@

@@

\caption{\label{fig-cj16}Comparative judgment model, SCM, probabilistic
and statistical model assuming different discriminal dispersions for the
student's traits}

@@

Third, we define \(T_{IA}\) as a linear combination of the students'
underlying writing-quality traits \(T_{I}\), the effects of relevant
text-related variables \(\beta_{XA}X_{IA}\) (e.g., text length on
quality), and the texts' idiosyncratic errors \(e_{IA}\). We assume that
each element of \(e_{IA}\) is independent and identically distributed
\((iid)\) according to a Normal distribution, with a mean of zero and a
standard deviation \(p_{IA}\). The parameter \(p_{IA}\) is defined as a
proportion of 1 to establish the scale of the latent trait, which is
required in any latent variable model \citep{Depaoli_2021}. Note that
because \(e_{IA}\) follows a Normal distribution, \(T_{IA}\) is also
normally distributed. This result indicates that our model reverts to
the original assumptions for the discriminal processes of stimuli
proposed by \citet{Thurstone_1927b} (see
Table~\ref{tbl-thurstone_cases}).

Fourth, we define \(T_{I}\) as a linear combination of the effects of
relevant student-related variables, \(\beta_{XI} X_{I}\), and students'
idiosyncratic errors, \(e_{I}\). As before, we assume that each element
of \(e_{I}\) is \(iid\) Normal with a mean of zero. However, we consider
the standard deviation, \(s_{XI}\), varies depending on the teaching
method group to which each student belongs. Given the example in
Section~\ref{sec-theoretical}, where the teaching method
\(X_{I} = 1,2\), the model sets the following constraint to anchor the
scale of the latent trait \citep{Depaoli_2021}:
\(\sum_{g=1}^{2} s_{XI}[g]/2 = 1\). Note that this parametrization of
\(e_{I}\) relaxes the equal dispersion assumption for the stimuli,
effectively addressing the concerns raised in
Section~\ref{sec-theory-issue1a}.

Fifth, we define \(B_{JK}\) as a linear combination of the judges'
individual bias \(B_{J}\), the effects of relevant judgment-related
variables \(\beta_{ZK}Z_{JK}\) (e.g., how the number of judgments a
judge makes affect the assessment of quality), and judgment-specific
idiosyncratic errors \(e_{JK}\). We assume that each element of
\(e_{JK}\) is \(iid\) Normal with a mean of zero and a standard
deviation \(p_{JK}\), defined as a proportion of \(1\), to anchor the
scale of the latent trait \citep{Depaoli_2021}.

Sixth, we define \(B_{J}\) as a linear combination of the effects of
relevant judge-level variables \(\beta_{ZJ}Z_{J}\) (e.g., judgment
expertise) and judge-specific idiosyncratic errors \(e_{J}\). Each
element of \(e_{J}\) is likewise assumed to be \(iid\) Normal with mean
zero and a standard deviation \(s_{ZJ}\) that depends on the groups to
which each judge belongs. For instance, if \(Z_{J} = 1, \dots, 3\)
represents three groups of judges with varying expertise, the model
imposes the constraint \(\sum_{g=1}^{3} s_{ZJ}[g]/3 = 1\) to fix the
scale of the latent trait \citep{Depaoli_2021}.

Finally, to transform model~\ref{fig-cj16_stat} into a practical
statistical tool for analyzing paired comparison data, we use
\emph{Bayesian inference methods}. The \textbf{Declarations} section of
this document includes a link to the model, along with an alternative
specification that assumes equal discriminal dispersions. We implemented
both versions of the model using \texttt{Stan} \citep[version
\(2.26.1\),][]{Stan_2020}. Although the implementation was successful,
we do not report the results here, as they fall outside the scope of
this document.

@@

@@

\begin{equation}\phantomsection\label{eq-group_diff}{
E[Y_{i} \mid T_{i}=1] - E[Y_{i} \mid T_{i}=2]
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-ICE}{
\tau_{i} = Y_{i} \mid do(T_{i}=1) - Y_{i} \mid do(T_{i}=2)
}\end{equation}

where \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \mid do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \mid do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Here, an
\emph{intervention} involves assigning a constant value to the treatment
variable for each student's potential outcomes. Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address this challenge, the structural approach extends
the ICE to the \emph{average causal effect} (ACE,
Equation~\ref{eq-ACE}), representing the average difference between the
students' observed potential outcomes and their counterfactual
counterparts.

\begin{equation}\phantomsection\label{eq-ACE}{
\begin{aligned}
\tau & = E[\tau_{i}] \\
  & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)]
\end{aligned}
}\end{equation}

Even though counterfactuals are unobservable, researchers can still
identify the ACE from associational estimates by leveraging the
structural approach. The approach identifies the ACE by statistically
conditioning data on a \emph{sufficient adjustment set} of variables
\(X\) \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones. When such a set
exists, then \(T\) and \(Y\) are \emph{d-separated} by \(X\)
(\(T \:\bot\:Y \mid X\)) \citep{Pearl_2009}, and \(X\) satisfies the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}. Here,
\emph{conditioning} describes the process of restricting the focus to
the subset of the population defined by the conditioning variable
\citep[pp.~32]{Neal_2020} (see Equation~\ref{eq-CACE}).

@@

\begin{equation}\phantomsection\label{eq-CACE}{
CACE_{t} = E[Y_{i} \mid T_{i}=t, X]
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mACE1}{
\begin{aligned}
  \tau & = E[Y_{i} \mid do(T_{i}=1)]- E[Y_{i} \mid do(T_{i}=2)] \\
  & = E_{X}[CACE_{1} - CACE_{2}] \\
  & = E_{X}\left[ E[Y_{i} \mid T_{i}=1, X] - E[Y_{i} \mid T_{i}=2, X] \right]
\end{aligned}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-prop_rule}{
P(\theta \mid X_{1}, X_{2}, \dots, X_{P}) \propto P(X_{1}, X_{2}, \dots, X_{P}\mid \theta) \cdot P(\theta)
}\end{equation}
